---
title: "Exploratory Analysis"
output: github_document
---

```{r setup}
library(tidyverse)
library(patchwork)
```

Global settings for themes (for figures).
```{r}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,          #aspect ratio: height over width
  out.width = "90%"      # figure doesn't go all the way out on full(100%) page
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis", # to override ggplot default colors.
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Load NOAA weather data, and add month variable.
```{r}
weather_df =  
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>% #package to add month var
  select(name, id, everything())
```

group_by name and month
```{r}
weather_df %>%
  group_by(name, month)
```

Count number of observations in each month.
```{r}
weather_df %>%
  group_by(month) %>%
  summarize(n_obs = n())
```

Group and summarize by two variables
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(n_obs = n())
```
Or, use "count".
```{r}
weather_df %>%
  count(month, name = "n_obs")
```

Summarize: number of observations in each month and number of distinct values of date in each month(e.g. feb = will have 28 days).
```{r}
weather_df %>%
  group_by(month) %>%
  summarize(
    n_obs = n(),
    n_days = n_distinct(date))
```

2 x 2 tables: tidy
```{r}
weather_df %>% 
  mutate(
    cold = case_when(
      tmax <  5 ~ "cold",
      tmax >= 5 ~ "not_cold",
      TRUE      ~ ""
  )) %>% 
  filter(name != "Waikiki_HA") %>% 
  group_by(name, cold) %>% 
  summarize(count = n())
```

reorganize to standard (untidy)table, using pivot_wider or janitor::tabyl
```{r}
weather_df %>% 
  mutate(cold = case_when(
    tmax <  5 ~ "cold",
    tmax >= 5 ~ "not_cold",
    TRUE     ~ ""
  )) %>% 
  filter(name != "Waikiki_HA") %>% 
  janitor::tabyl(name, cold)
```

General summaries
```{r}
weather_df %>%
  group_by(month) %>%
  summarize(
    mean_tmax = mean(tmax),
    mean_prec = mean(prcp, na.rm = TRUE),
    median_tmax = median(tmax),
    sd_tmax = sd(tmax))
```
Summaries with 2 variables.
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(
    mean_tmax = mean(tmax),
    median_tmax = median(tmax))
```

Summarize multiple columns with "across".
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(across(tmin:prcp, mean))
```

Creat a plot based on monthly summary.
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(mean_tmax = mean(tmax)) %>%
  ggplot(aes(x = month, y = mean_tmax, color = name)) + 
    geom_point() + geom_line() + 
    theme(legend.position = "bottom")
```

"Mutate" with "group_by": e.g compare daily max temp to annual average max temp
```{r}
weather_df %>%
  group_by(name) %>%
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) + 
    geom_point() 
```

Window functions: Ranking
```{r}
weather_df %>%
  group_by(name, month) %>%
  mutate(temp_ranking = min_rank(tmax)) # to get max temp within month

#day with the lowest max temperature within each month:
weather_df %>%
  group_by(name, month) %>%
  filter(min_rank(tmax) < 2)

#three days with the highest max temperature:
weather_df %>%
  group_by(name, month) %>%
  filter(min_rank(desc(tmax)) < 4)
```

Offsets:e.g. to find the day-by-day change in max temp within each station over the year.
```{r}
weather_df %>%
  group_by(name) %>%
  mutate(temp_change = tmax - lag(tmax)) # gives temp change from previous day

# to quantify the day-by-day variability in max temperature, or to identify the largest one-day increase:
weather_df %>%
  group_by(name) %>%
  mutate(temp_change = tmax - lag(tmax)) %>%
  summarize(
    temp_change_sd = sd(temp_change, na.rm = TRUE),
    temp_change_max = max(temp_change, na.rm = TRUE))
```

Learning Assessment:Import, clean, and summarize the PULSE data to examine the mean and median at each visit. Export the results of this in a reader-friendly format.
```{r}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") %>%
  select(id, visit, everything()) %>%
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) %>%
  arrange(id, visit)

pulse_data %>% 
  group_by(visit) %>% 
  summarize(
    mean_bdi = mean(bdi, na.rm = TRUE),
    median_bdi = median(bdi, na.rm = TRUE)) %>% 
  knitr::kable(digits = 3)
```
End assessment ------------

